
# preliminary proteomics tool research

msn base:
https://bioconductor.org/packages/release/bioc/html/MSnbase.html

AIO solution for general proteomics processing from the mass spectometry to visualisation

msgf+:
for protein identification
e value calculation
minimal as possible protein clusters
protdb or smth files to compare to


# dividing of tasks
me: confusion matrix
jarno: pls-da
yamila: normalisation & imputation
ramon: pca





# confusion matrix
what do i need for a confusion matrix and what is it for?:

I need candidates from the pca for biomarker, since this is proteiomics I need proteins.
I also need a threshold to determine what is high and what is low expression

the matrix is for finding the precision of potential biomarkers(prediction data) in comparison to the biliary viability groups(actual data)




loading in count data
```{r}
load("/students/2024-2025/Thema08/liver-transplant/proteomics/count_data.Rdata")
```

loading in meta data
```{r}
load("/students/2024-2025/Thema08/liver-transplant/proteomics/meta_data.Rdata")

```

loading in imputed count data
```{r}

load("/students/2024-2025/Thema08/liver-transplant/proteomics/normalized_imputed_count_data.Rdata")

```

seperate into 2 groups
```{r}
high_viability_data <- assay_norm[, grep("^High", colnames(assay_norm))]
low_viability_data <- assay_norm[, grep("^Low", colnames(assay_norm))]


```

```{r}
type(assay_norm)
```
```{r}
colnames(assay_norm)
```

summary
```{r}

summary_high_vial <- DataFrame(high_viability_data)
summary(summary_high_vial)


```


protein to use as potential biomarker
```{r}
pot_biomarker <- "IGLV7-46"
```

```{r}
pot_bio_marker_high_counts <- DataFrame(high_viability_data[pot_biomarker, ])

```

biomarker corresponding count values
```{r}
pot_bio_marker_high_counts <- data.matrix(high_viability_data[pot_biomarker, ])
```

```{r}
pot_bio_marker_low_counts <- data.matrix(low_viability_data[pot_biomarker, ])
```


determining low/high expression categorisation (in this case I will use the median of the potential biomarker's expression values)

I am redoing the flow of this to optimise a little bit, because we might be doing this on a larger scale.

so instead of splitting the groups immediatly into low/high viability, I will first get the pot_biomarker's corresponding count val

```{r}
# clearing env
rm(list=ls())

# loading in imputed and normalised data
load("/students/2024-2025/Thema08/liver-transplant/proteomics/normalized_imputed_count_data.Rdata")

# choosing potential biomarker protein (current one is random)
pot_biomarker <- "IGLV7-46"

# filtering for the counts of only the potential biomarker
pot_bio_marker_counts <- data.matrix(assay_norm[pot_biomarker, ])

# determining the median for the expression categorisation threshhold
expression_categorisation_threshold <- median(pot_bio_marker_counts)

# divinding the counts into low and high viability groups
high_viability_pot_bio_marker_counts <- 
    data.matrix(pot_bio_marker_counts[grep("^High", rownames(pot_bio_marker_counts)),])

low_viability_pot_bio_marker_counts <- 
    data.matrix(pot_bio_marker_counts[grep("^Low", rownames(pot_bio_marker_counts)),])


```

making the matrixes into dataframes for filtering purposes
```{r}
high_viability_pot_bio_marker_counts_df <- data.frame(high_viability_pot_bio_marker_counts)
low_viability_pot_bio_marker_counts_df <- data.frame(low_viability_pot_bio_marker_counts)
```

checking on the df
```{r}
high_viability_pot_bio_marker_counts_df
```
load in dyplr for count function
```{r}
library("dyplr")
```
test on initial matrix value of top left aka high biliary and high expression
```{r}
true_true <- sum(high_viability_pot_bio_marker_counts_df$high_viability_pot_bio_marker_counts
 > expression_categorisation_threshold)
```

matrix values calculated (I have made it so expressions equal to the threshold still count as highly expressed, note that there is no logic behind this) 
```{r}
# confusion matrix scores

# biliary viability is high and so is protein count
true_true <- 
    sum(high_viability_pot_bio_marker_counts_df$high_viability_pot_bio_marker_counts
 >= expression_categorisation_threshold)

# biliary viability is low and protein count is high
false_true <- 
    sum(low_viability_pot_bio_marker_counts_df$low_viability_pot_bio_marker_counts
 >= expression_categorisation_threshold)

# biliary viability is low and so is protein count
true_false<- 
    sum(low_viability_pot_bio_marker_counts_df$low_viability_pot_bio_marker_counts
 < expression_categorisation_threshold)

# biliary viability is high and protein count is low
false_false <- 
    sum(high_viability_pot_bio_marker_counts_df$high_viability_pot_bio_marker_counts
 < expression_categorisation_threshold)

```

loading in the caret library for automatic confusion matrices
```{r}
if (!require("caret", quietly = TRUE)) {
  install.packages("caret")
  library("caret")
}
```

inputting the values into the ConfusionMatrix() funtion. reference: https://www.rdocumentation.org/packages/caret/versions/7.0-1/topics/confusionMatrix
```{r}
confusionMatrix(  
    # transform into a table
    as.table(matrix(c(false_false, false_true, false_false, true_true), nrow = 2)),
  positive = "positive"
)
```
I think it doesn't work because the data is in an incompatible format

put the data in a matrix instead of a table to prep it for caret
```{r}
matrix(c(true_false, false_true, false_false, true_true),
    nrow = 2,
    dimnames = list(
    protein_expression = c("low", "high"),
    biliary_viability = c("low", "high")))
```
bit of an unclear confusion matrix compared to the usual format

make the same matrix but with labels that are more standard
```{r}
conf_matrix <- matrix(c(true_false, false_true, false_false, true_true),
    nrow = 2,
    dimnames = list(
    protein_expression = c("false", "true"),
    biliary_viability = c("false", "true")))
```
due to this njot working in the caret function it could still make a good backup handmade matrix

better layout(subjectiovely speaking) matrix
```{r}
conf_matrix <- matrix(
  c(true_true, false_false, 
    false_true, true_false),
  nrow = 2,
  dimnames = list(
    protein_expression = c("high", "low"),      
    biliary_viability = c("high", "low")    
  )
)
```


put it in caret function again
```{r}
confusionMatrix(conf_matrix,
                positive = "positive") # seems to only be lable according to the documentation
```
according to the documentation it has to be of "class table" so back to table

```{r}
confusionMatrix(as.table(conf_matrix), #data
                positive="high")
```
subscript out of bounds again


turns out the positive wasn't a lable but had to instead match the lable given to the true option in the table object
```{r}
confusionMatrix(as.table(conf_matrix), #data
                positive="true")
```


The initial loading in of the Rdata doesn't work anymore because the file cannot be found for some reason.
So I will try to navigate to the folder before once again loading in the Rdata object
```{r}
# navigating to the folder the file is in
setwd("/students/2024-2025/Thema08/liver-transplant/proteomics/")

# checking on the current directory
getwd()

# load in the file
load("normalized_imputed_count_data.Rdata")

```
This still doesn't work

turns out it's because the file got renamed so we change the loading in chunk to the following:
```{r}
load("/students/2024-2025/Thema08/liver-transplant/proteomics/normalized_imputed_count_data_30min.Rdata")

```

I want to make this script more efficent as we might (hopefully) need to upscale this.

## final version of confusion matrix script

clear environment of previous objects
```{r}
# clear env
rm(list=ls())
```

load in caret library
```{r}
if (!require("caret", quietly = TRUE)) {
  install.packages("caret")
  library("caret")
}
```

load in data
```{r}
#oad in data
load("/students/2024-2025/Thema08/liver-transplant/proteomics/normalized_imputed_count_data_30min.Rdata")
```

define potential bio marker protein
```{r}
# potential biomarker definitin
pot_biomarker <- "IGLV7-46"
```

extract only relevant expression data
```{r}
# filter for expression values of only the biomarker
pot_bio_marker_counts <- assay_norm[pot_biomarker, ]
```

threshold definition
```{r}
# get median threshhold (note that this is temp and if another is neccesary then we will do that)
expression_threshold <- median(pot_bio_marker_counts)
```

split the expression data into high and low viability groups with grepl instead of grep so I won't have to run it twice by basically creating a mask for both viabilities whilst keeping the original pot_bio_marker_counts
```{r}
# split into viability groups
high_viability <- grepl("^High", names(pot_bio_marker_counts))
low_viability <- !high_viability
```

"calculate" matrix components. this time with better lables cause I found the previous ones confusing
```{r}
true_positive <- sum(pot_bio_marker_counts[high_viability] >= expression_threshold)
false_positive <- sum(pot_bio_marker_counts[low_viability] >= expression_threshold)
true_negative <- sum(pot_bio_marker_counts[low_viability] < expression_threshold)
false_negative <- sum(pot_bio_marker_counts[high_viability] < expression_threshold)
```

make a matrix with the elements generated. honestly it would be more efficient to make a table directly but that's a format that offers less control and the effort required for it doesn't seem worth it.
```{r}
conf_matrix <- matrix(
  c(true_positive, false_negative, # fuckass layout for the actual data in the matrix
    false_positive, true_negative),
  nrow = 2, # amount of rows
  dimnames = list(
    protein_expression = c("high", "low"),  # labels for the rows
    biliary_viability = c("high", "low")    # labels for the columns
  )
)
```

convert to confusionMatrix object of caret and all the stats that accompany it
```{r}
final_matrix <- confusionMatrix(
  as.table(conf_matrix), # convert matrix obj to table so it's suited in terms of input
  positive = "high" # give the label for the equivalent of positive in a confusion matrix
)

print(final_matrix)
```


### conclusion on confusion matrix
basically for efficiency's sake I tried to avoid using complecated data structures like matrices and dataframes. and lessen the amount of casting I used for convenience
