---
title: "Logbook Omics liver transplant"
author: "Jarno Jacob Duiker"
date: "2025-04-28"
output: html_document
    toc: true                    # Table of Contents
    toc_float: true              # Floating TOC
    code_folding: show           # Toggle code chunks
    df_print: paged
bibliography: bibliography.bib    # Path to your .bib file
---

### Introduction

This is a logbook for the Omics liver transplant project. The goal of this project is to perform a PCA on both the proteomics data and RNA transcriptomics data, and find a link.

### Load libraries

```{r}
```

## Reading the Paper

The paper selected for this multi-omics project is the *Bile proteome reveals biliary regeneration during normothermic preservation of human donor livers* @thorne2023

The paper looks at the effects of NMP (Normothermic machine perfusion) on livers that are going to be transplanted that might not be in the best shape, meaning they have high biliary injury's (\

## Proteomics data

```{r}
data_proteomics <- read.delim("data/NMP_Bile_Proteomics_Report.txt", sep = "\t")
data.frame(data_proteomics)
head(data_proteomics)

```

```{r}
colnames_proteomics <- names(data_proteomics)
colnames_proteomics
```

```{r}
summary(data_proteomics)
```

x
