---
title: "Untitled"
author: "Yamila Timmer"
date: "2025-05-08"
output: html_document
---

```{r}
# Load packages
library(pathview)
library(tidyverse)
library(readxl)
library(DEP)
library(SummarizedExperiment)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(clusterProfiler)
```

# 7/8/9 may

- Looking into what data we need and how to process this in R

Report contains all the quantified proteins (results from protein identification/quantification), is a tsv file and can be read with read.delim. In this file
rows is proteins/genes columns is sample id.

```{r}
data <- read.delim("../../../NMP_Bile_Proteomics_Report.txt", header = TRUE)

```

File below contains metadata from transcriptomics. Source name refers to the FASTA ID's (transcriptomics), characteristics individual refers to liver number (I think?), possible interesting metadata to keep: age, sex,
```{r}
meta_data <- read.delim("../../../E-MTAB-13501.sdrf.txt")

```

Idea: compare between biliary viability score group (high/low) per time point, so at 30 min, 150 min and end.

Files below are supp data 3a + 4a and seems to contain transcriptomics data (gene expression)
```{r}
sup_data <- read_excel("../../../supp_data.xlsx", sheet = "Figure 3a", col_names = TRUE, skip = 4, trim_ws = TRUE) # skip lines so the empty excel lines are not left in
```


```{r}
sup_data2 <- read_excel("../../../supp_data.xlsx", sheet = "Figure 4a", col_names = TRUE, skip = 3, trim_ws = TRUE) # skip lines so the empty excel lines are not left in

```

Make dataframe for identifying all sample no:
- From the regular data (NMP_Bile_Proteomics_Report.txt) it is unclear which sample number belongs to which liver/measure point
- Dataframe shows sample ID, liver number (so what liver the sample was derived from) and timepoint from when sample was taken (after 30 min/150 min/end)

# Creating Metadata Dataframe
```{r}
sample_id1 <- sup_data[c("Sample", "Liver number", "Timepoint")]

sample_id2 <- sup_data2[c("Sample", "Biliary viability score group", "Total BDI score group", "Transplant")]

# Only merges when sample ID is in both dataframes, leaves some livers out as there seems to be some missing data?
merged_sample_id <- inner_join(sample_id1, sample_id2, by = "Sample")

```


# Pre-processing steps

Change colnames to NMP_Bile_Proteomics_X, in order to be able to filter on samples later
```{r}
colnames(data)[4:ncol(data)] <- sub(".*?(NMP_Bile_Proteomics_[0-9]+).*", "\\1", colnames(data)[4:ncol(data)]) # regex that removes everthing, but the "NMP_Bile_Proteomics_X" identifier and replaces colnames of data with this, we need that to match it to the metadata
```


```{r}
# adds unique identifier columns based on name/id columns, new columns are added to 'data_unique' as 'name' and

data_unique <- make_unique(data, 
                           names = "PG.Genes", 
                           ids = "PG.ProteinAccessions", 
                           delim = ";")
```

```{r}
# Filter only the 30 min samples
samples_30min <- merged_sample_id$Sample[merged_sample_id$Timepoint == "30min"]
cols_30min <- c("ID", "name", "PG.ProteinGroups", "PG.Genes","PG.ProteinAccessions", samples_30min)

# Filter only the 30 min samples
samples_150min <- merged_sample_id$Sample[merged_sample_id$Timepoint == "150min"]
cols_150min <- c("ID", "name", "PG.ProteinGroups", "PG.Genes","PG.ProteinAccessions", samples_150min)

samples_30min_data <- data_unique[, colnames(data_unique) %in% cols_30min]
samples_30min_metadata <- merged_sample_id[merged_sample_id$Sample %in% samples_30min, ]

samples_150min_data <- data_unique[, colnames(data_unique) %in% cols_150min]
samples_150min_metadata <- merged_sample_id[merged_sample_id$Sample %in% samples_150min, ]
```


# Creating Summarized Experiment Object

```{r}
sample_columns_30min <- grep("NMP.", colnames(samples_30min_data))

experimental_design_30min <- data.frame(
    label = colnames(samples_30min_data[, sample_columns_30min]), # has to match colnames of samples 
    condition = samples_30min_metadata$`Biliary viability score group`,
    replicate = colnames(samples_30min_data[, sample_columns_30min]) # there are no replicates in this case, but they still need a unique identifier because duplicate rownames are not allowed
)

sample_columns_150min <- grep("NMP.", colnames(samples_150min_data))

experimental_design_150min <- data.frame(
    label = colnames(samples_150min_data[, sample_columns_150min]),  #has to match colnames of samples 
    condition = samples_150min_metadata$`Biliary viability score group`,
    replicate = colnames(samples_150min_data[, sample_columns_150min]) # there are no replicates in this case, but they still need a unique identifier because duplicate rownames are not allowed
)

```

```{r}
# Convert NaN to NA, which is neccessary to create SE object and for normalisation
samples_30min_data <- samples_30min_data %>% mutate_all(~ifelse(is.nan(.), NA, .))
samples_150min_data <- samples_150min_data %>% mutate_all(~ifelse(is.nan(.), NA, .))

load("../../top_x_genes_150_new.Rdata")

# Select only rows with 200 genes with highest loadings
top_200_genes <- top_x_genes_150[1:80]
#top_200_genes
samples_150min_data <- samples_150min_data%>%
  filter(samples_150min_data$PG.Genes %in% top_200_genes)

```

```{r}

data_se_30min <- make_se(samples_30min_data, sample_columns_30min, experimental_design_30min)

data_se_150min <- make_se(samples_150min_data, sample_columns_150min, experimental_design_150min)

```

# Removing Samples/Proteins with high missing values
```{r}
mean_missing_percentage_samples_30min <- median(colMeans(is.na(assay(data_se_30min))) * 100)
mean_missing_percentage_samples_30min

mean_missing_percentage_samples_150min <- median(colMeans(is.na(assay(data_se_150min))) * 100)
mean_missing_percentage_samples_150min

# bad_samples_30min <- names(missing_per_sample_30min[missing_per_sample_30min > 70])
# 
# 
# # samples with >70% missing values, would rely too much on imputation
# 
# 
# bad_samples_30min <- names(mean_missing_percentage_samples_30min
#                            [mean_missing_percentage_samples_30min > 70])
# 
# data_se_30min <- data_se_30min[, !colnames(data_se_30min) %in% bad_samples_30min]
# 
# missing_per_sample_150min <- colMeans(is.na(assay(data_se_30min))) * 100
# high_amount_missing_sample_150min <- missing_per_sample_150min
# bad_samples_150min <- names(mean_missing_percentage_samples_150min
#                             [mean_missing_percentage_samples_150min > 70])
# 
# data_se_150min <- data_se_150min[, !colnames(data_se_150min) %in% bad_samples_150min]
```

```{r}
# Calculate the median average presence of proteins per sample, in the article this was 41.5%
mean_protein_presence_30min <-  median(rowMeans(!is.na(assay(data_se_30min))))
mean_protein_presence_30min

mean_protein_presence_150min <- median(rowMeans(!is.na(assay(data_se_150min))))
mean_protein_presence_150min
```

```{r}
# Keep all proteins with at least 30% presence across all samples (based on mean_protein_presence)

data_se_30min_filter <- filter_proteins(data_se_30min, type = "condition", thr = 2)

data_se_150min_filter <- filter_proteins(data_se_150min, type = "condition", thr = 2)
```


# Normalisation

```{r}
data_norm_30min <- normalize_vsn(data_se_30min_filter)
data_norm_150min <- normalize_vsn(data_se_150min_filter)

```

```{r}

meanSdPlot(data_norm_30min)
meanSdPlot(data_norm_150min)

```
```{r}
plot_frequency(data_norm_30min)
plot_frequency(data_norm_150min)
```
```{r}
plot_detect(data_norm_30min)
plot_detect(data_norm_150min)

```

On average the missing values come from proteins with low intensities (points to MNAR), an imputation method that is suited for this is (“QRILC”) quantile regression-based left-censored function

After imputation with QRILC, a second "top" exists, with low intensity, this is because it fills NA values with low intensity values that were failed to be measured in the first place (MNAR)



# Imputation
```{r}
#set.seed()
data_imp_30min <- impute(data_norm_30min, fun="QRILC")
data_imp_150min <- impute(data_norm_150min, fun="QRILC")

```

```{r}
plot_pca(data_imp_150min, indicate = "condition", n=71)

#plot_pca(data_imp_30min, indicate = "condition")
```


```{r}
plot_imputation(data_norm_150min, data_imp_150min)
plot_imputation(data_norm_30min, data_imp_30min)


```

# Differential Analysis (comparing groups)
```{r}
data_diff_30min <- test_diff(data_imp_30min, type = "all")
data_diff_150min <- test_diff(data_imp_150min, type = "all")


dep_30min <- add_rejections(data_diff_30min, alpha = 0.05, lfc = 1)
dep_150min <- add_rejections(data_diff_150min, alpha = 0.05, lfc = 1)

```

```{r}
plot_volcano(dep_30min, contrast = "Low_vs_High", label_size = 2, add_names = TRUE)
plot_volcano(dep_150min, contrast = "Low_vs_High", label_size = 2, add_names = TRUE)

```
```{r}
results_df_150min <- get_results(dep_150min)

```


# Pathway Analysis
## Subset on significant DEG's
```{r}
# Subset results for comparison, based on p-adjusted of 0.05 and/or below
low_vs_high_150min <- results_df_150min[results_df_150min$Low_vs_High_p.adj <= 0.05, ]
```

```{r}
# Create dataframe with corresponding entrez ID's for gene symbols
entrez_ID <- function(sig_df) {
    gene_names <- sig_df$name
    genes_entrez <- AnnotationDbi::select(org.Hs.eg.db, keys = gene_names,
                                          columns = c('ENTREZID'), keytype = 'SYMBOL')
}

low_vs_high_150min_entrez <- entrez_ID(sig_df = low_vs_high_150min)

```

```{r}
# Make dataframe that shows all pathways that have one or more DEG
pathway_low_vs_high_150min <- enrichKEGG(gene =
                                             low_vs_high_150min_entrez$ENTREZID,
                                         organism = 'hsa')

pathway_low_vs_high_150min_res <- pathway_low_vs_high_150min@result

# Only keep significantly differentially expressed pathways
pathway_low_vs_high_150min_res <- subset(pathway_low_vs_high_150min_res, pvalue <=
                                             0.05 & pvalue < p.adjust)


```

```{r}
# Make dataframe that inclused log fold change (Low_vs_High_ratio) and entrez id per significantly differentially expressed gene
low_vs_high_150min_log <-
    data.frame(low_vs_high_150min$Low_vs_High_ratio)
rownames(low_vs_high_150min_log) <- low_vs_high_150min_entrez$ENTREZID
```


```{r}
# Create visualisations for each pathway with at least 1 DEG

pathview_visualise <- function(comparison, pathway_id){
    pathview(gene.data = comparison, # chosen comparison
             pathway.id = pathway_id, # hsa ID
             species = "hsa", # human sapiens
             low = list(gene="magenta"), # underexpressed = purple
             high = list(gene="green")) # overexpressed = green
}

# Generate all pathway visualisations per comparison
comparisons = c(pathway_low_vs_high_150min_res, pathway_low_vs_high_30min_res)

# for (comparison in comparisons){
#     for (pathway_id in pathway_low_vs_high_150min_res$ID){
#         pathview_visualise(comparison = comparison, pathway_id=pathway_id)
#     }
# }
```


https://www.bioconductor.org/packages/devel/bioc/vignettes/DEqMS/inst/doc/DEqMS-package-vignette.html
